# Chemometrics {#sec-chemometrics}

In this section, we introduce a modeling paradigm commonly referred to as **Chemometrics**. According to [Wikipedia](https://en.wikipedia.org/wiki/Chemometrics),

> **Chemometrics** is the science of extracting information from chemical systems by data-driven means. Chemometrics is inherently interdisciplinary, using methods frequently employed in core data-analytic disciplines such as multivariate statistics, applied mathematics, and computer science, in order to address problems in chemistry, biochemistry, medicine, biology and chemical engineering. In this way, it mirrors other interdisciplinary fields, such as psychometrics and econometrics.

Chemometrics has been widely employed in soil spectroscopy, especially with the use of traditional preprocessing tools and Partial Least Squares Regression (PLSR). PLSR is a classic algorithm that is able to deal with the multivariate and multicollinear nature of spectra to create robust regression models. Along with regression, many other features have been built around PLSR and PCA that aid in the interpretation of models and prediction results. This section will present and discuss these possibilities.

The choice between classic chemometrics tools and those presented in the [**Processing**](#subsec-preparing) and [**Machine Learning**](#subsec-ml) sections will largely depend on the project application and familiarity with each framework. There is no perfect method, so we usually recommend testing both and choosing the one that best suits your needs.

The **mdatools** R package is a collection of chemometrics tools that share a common interface. It contains a series of features for preprocessing, exploring, modeling, and interpreting multivariate spectral data.

The **mdatools** framework is presented in @Kucheryavskiy2020. For full documentation of the available features, please visit <https://mdatools.com/docs/index.html>.

A list of all required packages for this section is provided in the following code chunk:

```{r, libraries, message=FALSE, warning=FALSE}
library("tidyverse") # Regular data wrangling
library("mdatools") # Chemometrics
library("yardstick") # Additional performance metrics
library("qs") # Loading serialized files
```

## Preparation of files

Instead of using the `train.csv` and `test.csv` files from the **Processing** section, we are going to preprocess and model the same spectral data using a different set of tools. For this, we use the same Neospectra database that is part of the Open Soil Spectral Library (OSSL).

```{r files, message=FALSE, warning=FALSE}
## Internet configuration for downloading large datasets
options(timeout = 10000)

## Reading serialized files using qs::qread_url()
neospectra.soil <- qread_url("https://storage.googleapis.com/soilspec4gg-public/neospectra_soillab_v1.2.qs")
dim(neospectra.soil)

neospectra.site <- qread_url("https://storage.googleapis.com/soilspec4gg-public/neospectra_soilsite_v1.2.qs")
dim(neospectra.site)

neospectra.nir <- qread_url("https://storage.googleapis.com/soilspec4gg-public/neospectra_nir_v1.2.qs")
dim(neospectra.nir)
```

From the original files, we select only the relevant data for modeling.

```{r preparation, message=FALSE, warning=FALSE}
# Selecting relevant site data
neospectra.site <- neospectra.site %>%
  select(id.sample_local_c, location.country_iso.3166_txt)

# Selecting relevant soil data
neospectra.soil <- neospectra.soil %>%
  select(id.sample_local_c, oc_usda.c729_w.pct)

# Selecting relevant NIR data and taking average across repeats
neospectra.nir <- neospectra.nir %>%
  select(id.sample_local_c, starts_with("scan_nir")) %>%
  group_by(id.sample_local_c) %>%
  summarise(across(everything(), mean)) %>%
  ungroup()

# Inspecting and renaming spectral column headers to numeric values
neospectra.nir %>%
  select(starts_with("scan_nir")) %>%
  names() %>%
  head()

old.names <- neospectra.nir %>%
  select(starts_with("scan_nir")) %>%
  names()

new.names <- gsub("scan_nir.|_ref", "", old.names)

neospectra.nir <- neospectra.nir %>%
  rename_with(~new.names, all_of(old.names))

spectral.column.names <- new.names

head(spectral.column.names)
tail(spectral.column.names)
```

We then join all three tables and remove samples without SOC values.

```{r join, message=FALSE, warning=FALSE}
# Joining data
neospectra <- left_join(neospectra.site,
                        neospectra.soil,
                        by = "id.sample_local_c") %>%
  left_join(., neospectra.nir, by = "id.sample_local_c")

# Filtering out samples without SOC values
neospectra <- neospectra %>%
  filter(!is.na(oc_usda.c729_w.pct))
```

A subset of the imported spectra can be visualized below.

```{r visualization, message=FALSE, warning=FALSE}
# Spectral visualization of a random subset
set.seed(42)
neospectra %>%
  sample_n(100) %>%
  pivot_longer(any_of(spectral.column.names),
               names_to = "wavelength",
               values_to = "reflectance") %>%
  mutate(wavelength = as.numeric(wavelength),
         reflectance = as.numeric(reflectance)) %>%
  ggplot(data = .) +
  geom_line(aes(x = wavelength, y = reflectance, group = id.sample_local_c),
            alpha = 0.25, linewidth = 0.25) +
  theme_light()
```

## Chemometric modeling

To start with chemometrics, **mdatools** offers a series of built-in functions for preprocessing spectra. We are going to combine **Savytzky-Golay first derivative** (`prep.savgol()`) with **Standard Normal Variate** (`prep.snv()`).

```{r preprocessing, message=FALSE, warning=FALSE}
# Preprocessing with SG 1st Der. and SNV
neospectra.prep <- neospectra %>%
  select(all_of(spectral.column.names)) %>%
  as.matrix() %>%
  prep.savgol(width = 11, porder = 1, dorder = 1) %>%
  prep.snv(.) %>%
  as_tibble() %>%
  bind_cols({neospectra %>%
      select(-all_of(spectral.column.names))}, .)

# Visualization of preprocessed spectra
set.seed(42)
neospectra.prep %>%
  sample_n(100) %>%
  pivot_longer(any_of(spectral.column.names),
               names_to = "wavelength",
               values_to = "reflectance") %>%
  mutate(wavelength = as.numeric(wavelength),
         reflectance = as.numeric(reflectance)) %>%
  ggplot(data = .) +
  geom_line(aes(x = wavelength, y = reflectance, group = id.sample_local_c),
            alpha = 0.25, linewidth = 0.25) +
  theme_light()
```

Before model calibration, we split the dataset into train and test sets using the same geographic rule employed in the **Processing** section — USA samples for training, African samples for testing.

```{r split, message=FALSE, warning=FALSE}
# Train split: USA samples
# Predictors and outcome are separated; log1p is applied to SOC to reduce skewness
neospectra.train <- neospectra.prep %>%
  filter(location.country_iso.3166_txt == "USA")

neospectra.train.predictors <- neospectra.train %>%
  select(all_of(spectral.column.names)) %>%
  as.matrix()

neospectra.train.outcome <- neospectra.train %>%
  select(oc_usda.c729_w.pct) %>%
  mutate(oc_usda.c729_w.pct = log1p(oc_usda.c729_w.pct)) %>%
  as.matrix()

# Test split: non-USA samples
neospectra.test <- neospectra.prep %>%
  filter(location.country_iso.3166_txt != "USA")

neospectra.test.predictors <- neospectra.test %>%
  select(all_of(spectral.column.names)) %>%
  as.matrix()

neospectra.test.outcome <- neospectra.test %>%
  select(oc_usda.c729_w.pct) %>%
  mutate(oc_usda.c729_w.pct = log1p(oc_usda.c729_w.pct)) %>%
  as.matrix()
```

We can now pass the preprocessed spectra to the PLSR algorithm.

PLSR compresses the spectra into several uncorrelated factors that simultaneously maximize covariance of the spectra and the soil property of interest — in this case, SOC. The resulting scores from these latent factors are then fed into a multivariate linear regression model.

::: {.callout-note}
## Note
If you want to learn more about PLSR and PCA, the **mdatools** documentation offers a good explanation. Similarly, take a look at this excellent chapter from [**All Models Are Wrong: Concepts of Statistical Learning**](https://allmodelsarewrong.github.io/pls.html).
:::

We are going to test up to 20 factors (`ncomp = 20`), run 10-fold cross-validation on the train data (`cv = 10`), center the spectra (no scale) before compression — performed globally across folds (`center = TRUE, scale = FALSE, cv.scope = 'global'`) — and use a data-driven method (`"ddmoments"`) to estimate critical limits for extreme and outlier observations. The model is named "SOC prediction model".

We can fit the PLSR by passing the train and test samples together in the call...

```{r plsr, message=FALSE, warning=FALSE}
set.seed(42)
pls.model.soc <- pls(x = neospectra.train.predictors,
                     y = neospectra.train.outcome,
                     ncomp = 20,
                     x.test = neospectra.test.predictors,
                     y.test = neospectra.test.outcome,
                     center = TRUE, scale = FALSE,
                     cv = 10, lim.type = "ddmoments", cv.scope = 'global',
                     info = "SOC prediction model")
```

Or run separately, which is useful when you want to apply a calibration model to new data independently.

```{r plsr_split, message=FALSE, warning=FALSE}
# Alternatively: fit calibration model first, then predict on test set
set.seed(42)
pls.model.soc.calibration <- pls(x = neospectra.train.predictors,
                                 y = neospectra.train.outcome,
                                 ncomp = 20,
                                 center = TRUE, scale = FALSE,
                                 cv = 10, lim.type = "ddmoments", cv.scope = 'global',
                                 info = "SOC prediction model")

pls.model.soc.predictions <- predict(pls.model.soc.calibration,
                                     x = neospectra.test.predictors,
                                     y = neospectra.test.outcome,
                                     cv = FALSE)
```

From these model fits, we can get a summary of the model.

```{r model_summary}
summary(pls.model.soc)
```

You can see that the number of selected components is 17, which was automatically chosen from the 10-fold CV results.

Let's plot the model. We will explore the visualization features in more detail soon.

```{r model_viz}
# Visualize representation limits, model coefficients, performance, and fit
plot(pls.model.soc, show.legend = TRUE)
```

There is a slight mismatch between CV and test performance, likely due to the geographical difference and underlying relationships (soil types, types of carbon) between the training (USA) and test (Africa) samples. Let's set the number of components that worked best for the test samples.

```{r select_ncomp}
# Performance is more consistent across train, CV, and test with 11 components
pls.model.soc <- selectCompNum(pls.model.soc, 11)
summary(pls.model.soc)
plot(pls.model.soc, show.legend = FALSE)
```

## Model and predictions interpretation

We can start the interpretation by looking at how observations are classified based on spectral dissimilarity. For this, **mdatools** uses both the **Q statistic** and **Hotelling $T^2$** as distance metrics for classifying samples as regular, extreme, or outlier.

The **Q statistic**, known as orthogonal distance in **mdatools**, measures the remaining spectral variance that is not accounted for during spectral compression and is not used by the model. Observations with unique features not well represented by the PLSR factors are flagged. The **Hotelling $T^2$**, on the other hand, focuses on the mean deviation of observation scores produced by the retained factors, flagging samples that are very distinct in score space. Together, these two complementary metrics can detect observations that are potentially underrepresented by the PLSR model. The classification thresholds are based on critical limits calculated from the calibration set; **mdatools** currently supports four methods, using the data-driven moments approach (`lim.type = "ddmoments"`) by default.

We can see that for the test set, several observations were flagged as potential extreme observations, with no spectral outliers detected.

```{r classification, message=FALSE, warning=FALSE}
# Sample categorization based on Q and T2 limits
outlier.detection <- categorize(pls.model.soc.calibration,
                                pls.model.soc.predictions,
                                ncomp = 11)
head(outlier.detection)

plotResiduals(pls.model.soc.predictions,
              cgroup = outlier.detection,
              ncomp = 11)
```

We can further explore the PLSR model with additional built-in plot functions — for example, examining model performance across the full range of components tested, for all data splits.

```{r performance_components}
plotRMSE(pls.model.soc, show.labels = TRUE)
```

With the predictions made, we can run a classic observed vs. predicted scatterplot.

```{r scatterplot}
# Predictions scatterplot
plotPredictions(pls.model.soc, show.line = TRUE, pch = 20, cex = 0.25)
abline(a = 0, b = 1)
```

We can also inspect the residuals.

```{r residuals}
# Inspection plot for outcome residuals
plotYResiduals(pls.model.soc, cex = 0.5, show.label = TRUE)
```

Another common visualization that aids model interpretation is variable importance. For PLSR, the Variable Importance in Projection (VIP) score is routinely employed.

```{r vip}
# Variable Importance in Projection (VIP) scores
plotVIPScores(pls.model.soc)
vip <- vipscores(pls.model.soc, ncomp = 11)
head(vip)
```

There are many other plot functions available for interpreting the PLSR model.

For example, how much of the original spectral variance (X variables) is retained by 11 components/factors? Around 90% for both train and test sets.

```{r x_variance}
# Cumulative variance retained from predictors (spectra)
plotXCumVariance(pls.model.soc, type = 'h', show.labels = TRUE, legend.position = 'bottomright')
```

As PLSR simultaneously maximizes variance in both the spectra (X) and the outcome (Y), how much of the outcome variance is retained by 11 components? About 70% in the train (= R^2) set and 88% in the test samples.

```{r y_variance}
# Cumulative variance retained from outcome
plotYCumVariance(pls.model.soc, type = 'b', show.labels = TRUE, legend.position = 'bottomright')
```

We can also visualize the scores of the latent factors. Here we plot the first three combinations.

```{r x_scores}
# Score plots for compressed spectra
plotXScores(pls.model.soc, comp = c(1, 2), show.legend = TRUE, cex = 0.5, legend.position = 'topleft')
# plotXScores(pls.model.soc, comp = c(1, 3), show.legend = TRUE, cex = 0.5, legend.position = 'bottomright')
```

During spectra compression, we can also visualize the estimated loadings for the first 5 components across the original spectral range.

```{r loadings}
# Loadings plot for the first 5 components
plotXLoadings(pls.model.soc, comp = c(1, 2, 3, 4, 5), type = 'l')
```

Lastly, we can explore all the internal information retained in the **pls** and **plsres** objects produced by **mdatools**.

```{r internal}
# Exploring internal info from model object
pls.model.soc$T2lim
pls.model.soc$Qlim

pls.model.soc$res$cal
pls.model.soc$res$cal$xdecomp

pls.model.soc$res$test
pls.model.soc$res$test$xdecomp
```

Let's extract the test predictions for a customized plot. Note that since we applied `log1p()` to the SOC outcome before modeling, we back-transform the predictions with `expm1()` to return to the original scale.

```{r predictions}
# Extracting test predictions
test.predictions <- pls.model.soc$res$test$y.pred
dim(test.predictions)

# Selecting predictions at component 11
test.predictions <- tibble(predicted = test.predictions[,11,1])

# Back-transforming predicted values from log1p to original scale
neospectra.test.results <- neospectra.test %>%
  select(-all_of(spectral.column.names)) %>%
  bind_cols(test.predictions) %>%
  rename(observed = oc_usda.c729_w.pct) %>%
  mutate(predicted = expm1(predicted))
```

We can then calculate additional evaluation metrics on the original scale.

```{r yardstick}
# Calculating performance metrics
neospectra.test.performance <- neospectra.test.results %>%
  summarise(n = n(),
            rmse = rmse_vec(truth = observed, estimate = predicted),
            bias = msd_vec(truth = observed, estimate = predicted),
            rsq = rsq_trad_vec(truth = observed, estimate = predicted),
            ccc = ccc_vec(truth = observed, estimate = predicted, bias = TRUE),
            rpd = rpd_vec(truth = observed, estimate = predicted),
            rpiq = rpiq_vec(truth = observed, estimate = predicted))

neospectra.test.performance
```

And plot the final observed vs. predicted scatterplot with **ggplot2**.

```{r ggplot}
# Final accuracy plot
performance.annotation <- paste0("Rsq = ", round(neospectra.test.performance[[1,"rsq"]], 2),
                                "\nRMSE = ", round(neospectra.test.performance[[1,"rmse"]], 2), " wt%")

p.final <- ggplot(neospectra.test.results) +
  geom_point(aes(x = observed, y = predicted)) +
  geom_abline(intercept = 0, slope = 1) +
  annotate(geom = "text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.2,
            label = performance.annotation, size = 3) +
  labs(title = "Soil Organic Carbon (wt%) test prediction",
       x = "Observed", y = "Predicted") +
  theme_light() + theme(legend.position = "bottom")

# Ensuring a square plot with equal axis limits
r.max <- max(layer_scales(p.final)$x$range$range)
r.min <- min(layer_scales(p.final)$x$range$range)

s.max <- max(layer_scales(p.final)$y$range$range)
s.min <- min(layer_scales(p.final)$y$range$range)

t.max <- round(max(r.max, s.max), 1)
t.min <- round(min(r.min, s.min), 1)

p.final <- p.final + coord_equal(xlim = c(t.min, t.max), ylim = c(t.min, t.max))
p.final
```
