# Machine Learning {#sec-ml}

In this section, we introduce the machine learning (ML) pipeline for estimating soil organic carbon (SOC) from NIR spectra using the [**tidymodels**](https://www.tidymodels.org/) framework. We start from the raw Neospectra database, apply prep preprocessing, compress the spectra with PCA, and then train and compare four models. We evaluate their generalization on African soil samples held out as a geographic test set.

The **tidymodels** framework is a collection of packages for modeling and machine learning that share a common design philosophy and interface. A comprehensive reference is available at the [tidymodels documentation](https://www.tidymodels.org/).

A list of all required packages for this section is provided in the following code chunk:

```{r libraries, message=FALSE, warning=FALSE}
library("tidyverse")
library("qs")         # Loading serialized files
library("prospectr")  # prep preprocessing
library("tidymodels")
library("probably")   # Conformal prediction intervals
library("rules")
library("Cubist")      # Cubist engine
library("kernlab")    # SVM engine
library("ranger")     # Random forest engine
library("yardstick")  # Performance metrics
tidymodels_prefer()   # Resolve common function name conflicts
```

## Data preparation {#subsec-preparing}

We use the NIR Neospectra Handheld database shared as part of the OSSL, which contains more than 2,000 unique soil samples scanned in replicates with different devices. The full pipeline covered here — download, filter, prep preprocessing, train/test split, PCA compression, and representation check — produces the feature matrix that feeds directly into the ML models below.

### Download and filter

```{r neospectra_read, eval=TRUE}
## Internet configuration for downloading large datasets
options(timeout = 10000)

## Reading serialized files using qs::qread_url()
neospectra.soil <- qread_url("https://storage.googleapis.com/soilspec4gg-public/neospectra_soillab_v1.2.qs")
dim(neospectra.soil)

neospectra.site <- qread_url("https://storage.googleapis.com/soilspec4gg-public/neospectra_soilsite_v1.2.qs")
dim(neospectra.site)

neospectra.nir <- qread_url("https://storage.googleapis.com/soilspec4gg-public/neospectra_nir_v1.2.qs")
dim(neospectra.nir)
```

We select only the relevant columns and average the replicate scans so each soil sample maps to a single spectrum. We also rename spectral columns to plain numeric wavelength values.

```{r neospectra_filter, eval=TRUE}
# How many samples from each country?
# Most samples are from USA, others from African countries
# We will use this column to define the train/test split
neospectra.site %>%
  count(location.country_iso.3166_txt)

# Selecting relevant site data
neospectra.site <- neospectra.site %>%
  select(id.sample_local_c, location.country_iso.3166_txt)

# Selecting relevant soil data
neospectra.soil <- neospectra.soil %>%
  select(id.sample_local_c, oc_usda.c729_w.pct)

# Selecting relevant NIR data and taking average across repeats
neospectra.nir <- neospectra.nir %>%
  select(id.sample_local_c, starts_with("scan_nir")) %>%
  group_by(id.sample_local_c) %>%
  summarise(across(everything(), mean)) %>%
  ungroup()

# Renaming spectral column headers to plain numeric wavelength values
old.names <- neospectra.nir %>%
  select(starts_with("scan_nir")) %>%
  names()

new.names <- gsub("scan_nir.|_ref", "", old.names)

neospectra.nir <- neospectra.nir %>%
  rename_with(~new.names, all_of(old.names))

spectral.column.names <- new.names

neospectra.nir[1:5, 1:5]
```

We then join all three tables and remove samples without SOC values.

```{r neospectra_join, eval=TRUE}
neospectra <- left_join(neospectra.site,
                        neospectra.soil,
                        by = "id.sample_local_c") %>%
  left_join(., neospectra.nir, by = "id.sample_local_c") %>%
  filter(!is.na(oc_usda.c729_w.pct))

neospectra[1:5, 1:5]
```

### Preprocessing

We combine **Savytzky-Golay first derivative** with **Standard Normal Variate** preprocessing — described in detail in the **Processing** section — to correct for offset and light scattering effects before compression.

```{r neospectra_prep, eval=TRUE}
neospectra.nir.prep <- neospectra %>%
  select(all_of(spectral.column.names)) %>%
  as.matrix() %>%
  savitzkyGolay(X = ., p = 2, w = 11, m = 1, delta.wav = 2) %>%
  prospectr::standardNormalVariate(X = .) %>%
  as_tibble() %>%
  bind_cols({neospectra %>%
      select(id.sample_local_c,
             location.country_iso.3166_txt,
             oc_usda.c729_w.pct)}, .)

# Quick visualization of preprocessed spectra
set.seed(42)
neospectra.nir.prep %>%
  sample_n(100) %>%
  pivot_longer(any_of(spectral.column.names),
               names_to = "wavelength",
               values_to = "reflectance") %>%
  mutate(wavelength = as.numeric(wavelength),
         reflectance = as.numeric(reflectance)) %>%
  ggplot(data = .) +
  geom_line(aes(x = wavelength, y = reflectance,
                group = id.sample_local_c),
            alpha = 0.25, linewidth = 0.25) +
  theme_light()
```

### Train/test split

We split by geography: USA samples for training, African samples for testing. This simulates a realistic scenario where a model trained on one region is applied to a new, geographically distinct region.

```{r neospectra_split, eval=TRUE}
neospectra.train <- neospectra.nir.prep %>%
  filter(location.country_iso.3166_txt == "USA")

neospectra.test <- neospectra.nir.prep %>%
  filter(location.country_iso.3166_txt != "USA")

nrow(neospectra.train)
nrow(neospectra.test)
```

### PCA compression and projection

**Partial Least Squares (PLS)** and **Principal Component Analysis (PCA)** are popular linear decomposition methods that help build regression models or explore the multivariate associations between spectra and variables of interest. Within this scope, the linear decomposition produces new orthogonal features such that the correlation between any pair of new variables is zero [@chang2001; @deSantana2023].

**PLS** is preferred when the goal is to find orthogonal features of the spectra that are most predictive of a soil property. On the other hand, if the goal is to find orthogonal features that maximally explain the variability of the spectra irrespective of the variable of interest, then **PCA** can be adopted.

Both methods reduce the dimensionality of the original data and produce uncorrelated features, controlling collinearity. A few orthogonal features obtained from the spectra can replace the original spectra matrix when training a model, with the downside of sacrificing some of the original spectral information.

We fit a PCA compression recipe using `step_normalize()` and `step_pca()` from **tidymodels**, retaining enough components to explain 99% of the cumulative variance (`threshold = 0.99`). The recipe is fitted (`prep()`) on the training data only, then applied to both sets with `bake()`.

```{r compression, eval=TRUE}
# Define and fit the PCA recipe on training data
pca.recipe <- recipe(neospectra.train) %>%
  update_role(id.sample_local_c, new_role = "id variable") %>%
  update_role(location.country_iso.3166_txt, new_role = "id variable") %>%
  update_role(oc_usda.c729_w.pct, new_role = "outcome") %>%
  update_role(any_of(spectral.column.names), new_role = "predictor") %>%
  step_normalize(all_predictors()) %>%
  step_pca(all_predictors(), threshold = 0.99, id = "pca")

pca.prep <- prep(pca.recipe, training = neospectra.train)

# How many components were retained?
pca.prep

# Explained variance per component from tidy()
pca.expvar <- tidy(pca.prep, id = "pca", type = "variance") %>%
  filter(terms == "percent variance") %>%
  pull(value)

# Extract training and test PC scores via bake()
train.pca.scores <- bake(pca.prep, new_data = neospectra.train)
test.pca.scores  <- bake(pca.prep, new_data = neospectra.test)

# PC column names produced by step_pca
pc.names <- train.pca.scores %>%
  select(starts_with("PC")) %>%
  names()

# PCA plot coloured by log1p(SOC)
p.pca <- ggplot(train.pca.scores) +
  geom_point(aes(x = PC01, y = PC02,
                 color = log1p(oc_usda.c729_w.pct)),
             alpha = 0.5, size = 0.5) +
  scale_colour_gradient(low = "gold", high = "darkred") +
  labs(title = "Neospectra PCA compression",
       x = paste0("PC1 (", round(pca.expvar[1], 2), "%)"),
       y = paste0("PC2 (", round(pca.expvar[2], 2), "%)")) +
  theme_light() +
  theme(legend.position = "bottom")

# Adding test samples to PCA plot
p.pca +
  geom_point(data = test.pca.scores,
             aes(x = PC01, y = PC02),
             size = 0.75) +
  labs(subtitle = "Black dots represent testing points")
```

### Representation flag

Before modeling, it is good practice to check whether the test samples are well represented by the training feature space. We use the **Q-statistic** (sum of squared residuals from PCA back-transformation) for this purpose. A good overview of this approach can be found in @deSantana2023.

Each spectrum is back-transformed from PC scores to the original spectral space using the loadings extracted from the prepped recipe via `tidy()`. The residual between the original and back-transformed spectrum is used to compute the Q-statistic, which is compared against a critical value derived from the training set at a 99% confidence level [@Jackson1979; @Laursen2011; @deSantana2023].

```{r backtransform, message=FALSE, error=FALSE}
# Extract loadings, centering means, and scaling SDs from the prepped recipe
n.pad <- nchar(length(pc.names))

pca.loadings <- tidy(pca.prep, id = "pca", type = "coef") %>%
  mutate(component = paste0("PC", formatC(as.integer(gsub("PC", "", component)),
                                          width = n.pad, flag = "0"))) %>%
  pivot_wider(names_from = component, values_from = value) %>%
  select(terms, all_of(pc.names)) %>%
  column_to_rownames("terms") %>%
  as.matrix()

norm.stats <- tidy(pca.prep, number = 1) %>%  # step_normalize is step 1
  select(terms, statistic, value) %>%
  pivot_wider(names_from = statistic, values_from = value)

train.center <- setNames(norm.stats$mean, norm.stats$terms)
train.scale  <- setNames(norm.stats$sd,   norm.stats$terms)

# Get PC scores as matrices (columns ordered to match loadings)
train.scores.mat <- train.pca.scores %>%
  select(all_of(pc.names)) %>%
  as.matrix()

test.scores.mat <- test.pca.scores %>%
  select(all_of(pc.names)) %>%
  as.matrix()

# Back-transform: rotate scores to spectral space, then rescale and recenter
train.bt <- train.scores.mat %*% t(pca.loadings)
test.bt  <- test.scores.mat  %*% t(pca.loadings)

train.bt <- sweep(train.bt, MARGIN = 2, FUN = "*", STATS = train.scale)
test.bt  <- sweep(test.bt,  MARGIN = 2, FUN = "*", STATS = train.scale)

train.bt <- sweep(train.bt, MARGIN = 2, FUN = "+", STATS = train.center)
test.bt  <- sweep(test.bt,  MARGIN = 2, FUN = "+", STATS = train.center)

# Original spectra matrices (rows ordered to match scores)
neospectra.train.spectra <- neospectra.train %>%
  select(any_of(spectral.column.names)) %>%
  as.matrix()

neospectra.test.spectra <- neospectra.test %>%
  select(any_of(spectral.column.names)) %>%
  as.matrix()

# Q-statistic: sum of squared differences between original and back-transformed
test.q.stats <- apply((neospectra.test.spectra - test.bt)^2,
                      MARGIN = 1, sum)
```

```{r q_critical, message=FALSE, error=FALSE}
# Critical value from the training set (1% significance level)
E     <- cov(neospectra.train.spectra - train.bt)
teta1 <- sum(diag(E))^1
teta2 <- sum(diag(E))^2
teta3 <- sum(diag(E))^3
h0    <- 1 - ((2*teta1*teta3) / (3*teta2^2))
Ca    <- 2.57 # 1% significance level
Qa    <- teta1*(1 - (teta2*h0*((1-h0)/teta1^2)) +
                  ((sqrt(Ca*(2*teta2*h0^2)))/teta1))^(1/h0)
Qa
```

```{r q_flag, message=FALSE, error=FALSE}
# Flag test samples and visualize on PCA plot
test.pca.scores <- test.pca.scores %>%
  mutate(q_stats = test.q.stats,
         represented = q_stats <= Qa)

p.pca +
  geom_point(data = test.pca.scores,
             aes(x = PC01, y = PC02, fill = represented),
             shape = 21, size = 1.5) +
  labs(fill = "Represented?")
```

## Machine learning with tidymodels

With the PC scores in hand, we now build and compare four predictive models. The `pc.names` vector defined during the compression step above identifies the predictor columns for all models.

### log1p transformation

SOC values have a right-skewed distribution. We apply `log1p()` to the outcome before modeling to reduce this skewness and improve model performance. Because the transformation is applied outside the recipe, we back-transform predictions with `expm1()` when reporting final results.

::: {.callout-note}
## Why log1p outside the recipe?
In tidymodels, `step_log()` applied to an outcome inside a recipe is skipped during prediction by default (`skip = TRUE`). This means predictions returned to the user are also on the log scale — you would still need to back-transform manually. Applying `log1p()` before training is more transparent and avoids confusion, especially for beginners.
:::

```{r log1p}
# Distribution of raw SOC values in the training set
hist(train.pca.scores$oc_usda.c729_w.pct, breaks = 50,
     main = "SOC distribution (raw)", xlab = "SOC (wt%)")

train.data <- train.pca.scores %>%
  mutate(oc_usda.c729_w.pct = log1p(oc_usda.c729_w.pct))

test.data <- test.pca.scores %>%
  mutate(oc_usda.c729_w.pct = log1p(oc_usda.c729_w.pct))

# Distribution after transformation
hist(train.data$oc_usda.c729_w.pct, breaks = 50,
     main = "SOC distribution (log1p)", xlab = "log1p(SOC)")
```

### Recipe

In tidymodels, a **recipe** defines the preprocessing steps applied to the data before modeling. Since we are working with PC scores that are already orthogonal, the main step here is normalization (centering and scaling) of the predictors. We assign roles to each column so that tidymodels handles non-predictor columns gracefully.

```{r recipe}
base.recipe <- train.data %>%
  recipe() %>%
  update_role(id.sample_local_c, new_role = "id variable") %>%
  update_role(location.country_iso.3166_txt, new_role = "id variable") %>%
  update_role(oc_usda.c729_w.pct, new_role = "outcome") %>%
  update_role(all_of(pc.names), new_role = "predictor") %>%
  step_normalize(all_predictors())

base.recipe
```

### Resamples

We define 10-fold cross-validation on the training data. These resamples are shared across all models to ensure a fair comparison. We set `save_workflow = TRUE` in the resampling control, which is required later for conformal prediction intervals via `int_conformal_cv()`.

```{r resamples}
set.seed(42)
train.folds <- vfold_cv(train.data, v = 10)

ctrl.resamples <- control_resamples(save_pred = TRUE,
                                    save_workflow = TRUE,
                                    extract = function(x) x)

ctrl.grid <- control_grid(save_pred = TRUE,
                          save_workflow = TRUE)
```

### Model specifications and hyperparameter grids

We compare four models that represent a range of complexity and interpretability:

- **Linear regression** — a simple baseline; with PC predictors it is conceptually analogous to PLSR.
- **Random forest** — an ensemble of decision trees, robust to outliers and nonlinear relationships.
- **Cubist** — a rule-based model combining decision trees with linear regression in the leaves, common in soil spectroscopy.
- **Support vector machine (SVM)** — a kernel-based model that works well in high-dimensional feature spaces.

```{r model_specs}
# Linear regression (no hyperparameters to tune)
spec.lm <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

# Random forest
spec.rf <- rand_forest(mtry  = tune(),
                       min_n = tune(),
                       trees = 500) %>%
  set_engine("ranger") %>%
  set_mode("regression")

# Cubist
spec.cubist <- cubist_rules(committees = tune(),
                            neighbors  = 0) %>%
  set_engine("Cubist") %>%
  set_mode("regression")

# Support vector machine (radial basis function kernel)
spec.svm <- svm_rbf(cost      = tune(),
                    rbf_sigma = tune()) %>%
  set_engine("kernlab") %>%
  set_mode("regression")
```

For models with hyperparameters, we define a regular grid to search over:

```{r grids}
grid.rf <- grid_regular(
  mtry(range = c(2L, floor(length(pc.names) / 3))),
  min_n(range = c(2L, 20L)),
  levels = 4)

grid.cubist <- grid_regular(
  committees(range = c(1L, 20L)),
  levels = 3)

grid.svm <- grid_regular(
  cost(range      = c(-1, 2)),
  rbf_sigma(range = c(-4, -1)),
  levels = 4)
```

### Workflows and tuning

Each model is wrapped in a **workflow** that combines the shared recipe with its model specification. We then run cross-validated hyperparameter optimization (HPO) for the three tunable models, and cross-validated resampling for the linear regression baseline.

```{r workflows_tuning, message=FALSE, warning=FALSE}
# Linear regression — fit_resamples (no HPO needed)
wf.lm <- workflow() %>%
  add_recipe(base.recipe) %>%
  add_model(spec.lm)

set.seed(42)
res.lm <- wf.lm %>%
  fit_resamples(resamples = train.folds,
                control = ctrl.resamples)

# Random forest — tune_grid
wf.rf <- workflow() %>%
  add_recipe(base.recipe) %>%
  add_model(spec.rf)

set.seed(42)
res.rf <- wf.rf %>%
  tune_grid(resamples = train.folds,
            grid = grid.rf,
            control = ctrl.grid)

# Cubist — tune_grid
wf.cubist <- workflow() %>%
  add_recipe(base.recipe) %>%
  add_model(spec.cubist)

set.seed(42)
res.cubist <- wf.cubist %>%
  tune_grid(resamples = train.folds,
            grid = grid.cubist,
            control = ctrl.grid)

# SVM — tune_grid
wf.svm <- workflow() %>%
  add_recipe(base.recipe) %>%
  add_model(spec.svm)

set.seed(42)
res.svm <- wf.svm %>%
  tune_grid(resamples = train.folds,
            grid = grid.svm,
            control = ctrl.grid)
```

### Selecting the best configuration per model

For each tunable model, we select the hyperparameter configuration that minimizes RMSE using the one-standard-error rule, which favors simpler models within one standard error of the best.

```{r select_best, message=FALSE, warning=FALSE}
best.rf <- select_by_one_std_err(res.rf,
                                 mtry, min_n,
                                 metric = "rmse")

best.rf

best.cubist <- select_by_one_std_err(res.cubist,
                                     committees,
                                     metric = "rmse")

best.cubist

best.svm <- select_by_one_std_err(res.svm,
                                  cost, rbf_sigma,
                                  metric = "rmse")

best.svm
```

We can compare cross-validation RMSE across all four models:

```{r compare_models, message=FALSE, warning=FALSE}
cv.metrics <- bind_rows(
  {res.lm %>%
    collect_metrics() %>%
    filter(.metric == "rmse") %>%
    summarise(model = "Linear Regression", mean_rmse = mean(mean))},
  {res.rf %>%
    collect_metrics() %>%
    filter(.metric == "rmse",
           mtry  == best.rf$mtry,
           min_n == best.rf$min_n) %>%
    summarise(model = "Random Forest", mean_rmse = mean(mean))},
  {res.cubist %>%
    collect_metrics() %>%
    filter(.metric == "rmse",
           committees == best.cubist$committees) %>%
    summarise(model = "Cubist", mean_rmse = mean(mean))},
  {res.svm %>%
    collect_metrics() %>%
    filter(.metric == "rmse",
           cost      == best.svm$cost,
           rbf_sigma == best.svm$rbf_sigma) %>%
    summarise(model = "Support Vector Machine", mean_rmse = mean(mean))})

cv.metrics %>% arrange(mean_rmse)
```

### Final model fit and test evaluation

We finalize each workflow with its best hyperparameter configuration and refit on the full training data. We then identify the best-performing model overall based on cross-validation RMSE and use it for final test set evaluation.

```{r final_fits, message=FALSE, warning=FALSE}
final.fit.lm <- wf.lm %>%
  fit(data = train.data)

final.fit.rf <- wf.rf %>%
  finalize_workflow(best.rf) %>%
  fit(data = train.data)

final.fit.cubist <- wf.cubist %>%
  finalize_workflow(best.cubist) %>%
  fit(data = train.data)

final.fit.svm <- wf.svm %>%
  finalize_workflow(best.svm) %>%
  fit(data = train.data)
```

Based on the CV comparison above, select the best model for test evaluation. In this example we use Cubist, but replace with whichever model had the lowest `mean_rmse` in your run:

```{r best_overall}
# Replace with the model that had the lowest CV RMSE in your run
best.final.fit <- final.fit.svm
best.res <- res.svm
best.params <- best.svm
```

We generate predictions on the test set and back-transform both observed and predicted values from log1p to the original SOC scale with `expm1()`.

```{r test_predictions, message=FALSE, warning=FALSE}
test.results <- predict(best.final.fit, new_data = test.data) %>%
  bind_cols(test.data %>%
              select(id.sample_local_c,
                     location.country_iso.3166_txt,
                     oc_usda.c729_w.pct)) %>%
  rename(predicted = .pred,
         observed  = oc_usda.c729_w.pct) %>%
  mutate(observed  = expm1(observed),
         predicted = expm1(predicted))

test.results[1:5,]
```

### Performance evaluation

We compute the same suite of performance metrics used in the **Chemometrics** section for direct comparability.

```{r performance, message=FALSE, warning=FALSE}
test.performance <- test.results %>%
  summarise(n    = n(),
            rmse = rmse_vec(truth = observed, estimate = predicted),
            bias = msd_vec(truth = observed, estimate = predicted),
            rsq  = rsq_trad_vec(truth = observed, estimate = predicted),
            ccc  = ccc_vec(truth = observed, estimate = predicted, bias = TRUE),
            rpd = rpiq_vec(truth = observed, estimate = predicted),
            rpiq = rpiq_vec(truth = observed, estimate = predicted))

test.performance
```

And plot the observed vs. predicted scatterplot.

```{r scatterplot, message=FALSE, warning=FALSE}
performance.annotation <- paste0("Rsq = ",  round(test.performance[[1, "rsq"]],  2),
                                 "\nRMSE = ", round(test.performance[[1, "rmse"]], 2), " wt%")

p.final <- ggplot(test.results) +
  geom_point(aes(x = observed, y = predicted)) +
  geom_abline(intercept = 0, slope = 1) +
  annotate(geom = "text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.2,
            label = performance.annotation, size = 3) +
  labs(title = "Soil Organic Carbon (wt%) — test prediction",
       x = "Observed", y = "Predicted") +
  theme_light()

r.max <- max(layer_scales(p.final)$x$range$range)
r.min <- min(layer_scales(p.final)$x$range$range)
s.max <- max(layer_scales(p.final)$y$range$range)
s.min <- min(layer_scales(p.final)$y$range$range)
t.max <- round(max(r.max, s.max), 1)
t.min <- round(min(r.min, s.min), 1)

p.final + coord_equal(xlim = c(t.min, t.max), ylim = c(t.min, t.max))
```

## Uncertainty quantification with conformal prediction

We use the **CV+ conformal inference** method from the **probably** package to compute prediction intervals. CV+ uses the residuals from each cross-validation fold to build a nonparametric reference distribution of errors, then applies a finite-sample correction to guarantee marginal coverage at the desired level (e.g., 95%).

::: {.callout-note}
## Note
`int_conformal_cv()` requires that the resampling results were produced with `save_workflow = TRUE` in the control object, which we set earlier. The function works directly on the tuning or resampling result object.
:::

```{r conformal, message=FALSE, warning=FALSE}
# Finalized workflow for the best model
final.wf.conformal <- wf.svm %>%
  finalize_workflow(best.svm)

# Resample with extract to populate .extracts for int_conformal_cv()
ctrl.conformal <- control_resamples(save_pred = TRUE,
                                    extract = function(x) x)

set.seed(42)
res.conformal <- final.wf.conformal %>%
  fit_resamples(resamples = train.folds,
                control   = ctrl.conformal)

# Build the conformal object from the CV resampling results of the best model
conformal.object <- int_conformal_cv(res.conformal)
conformal.object
```

We then compute 95% prediction intervals for the test set and back-transform everything to the original SOC scale.

```{r conformal_predict, message=FALSE, warning=FALSE}
# Prediction intervals on the log1p scale, then back-transformed
test.intervals <- predict(conformal.object,
                          new_data = test.data,
                          level    = 0.95) %>%
  bind_cols(test.data %>%
              select(id.sample_local_c,
                     oc_usda.c729_w.pct)) %>%
  rename(observed = oc_usda.c729_w.pct) %>%
  mutate(observed    = expm1(observed),
         .pred       = expm1(.pred),
         .pred_lower = expm1(.pred_lower),
         .pred_upper = expm1(.pred_upper),
         covered     = observed >= .pred_lower & observed <= .pred_upper)

test.intervals[1:5,]
```

We evaluate coverage (the fraction of test samples falling within their prediction interval) and mean interval width.

```{r conformal_coverage, message=FALSE, warning=FALSE}
# Coverage statistics
test.intervals %>%
  count(covered) %>%
  mutate(percentage = n / sum(n) * 100)

# Mean interval width
test.intervals %>%
  mutate(interval_width = .pred_upper - .pred_lower) %>%
  summarise(mean_width   = mean(interval_width),
            median_width = median(interval_width))
```

::: {.callout-note}
## On coverage and exchangeability
The theoretical guarantee of conformal prediction (marginal coverage ≥ 95%) holds under the **exchangeability** assumption — that training and test samples are drawn from the same distribution. In our case, the test set comes from a different geographical region (Africa vs. USA), which likely violates this assumption. Observed coverage below the nominal level is therefore expected and informative: it reflects the extent of distribution shift between the two sets.
:::

Finally, we visualize the prediction intervals alongside the observed values.

```{r conformal_plot, message=FALSE, warning=FALSE}
conformal.annotation <- paste0(
  "Rsq = ",       round(test.performance[[1, "rsq"]],  2),
  "\nRMSE = ",    round(test.performance[[1, "rmse"]], 2), " wt%",
  "\nCoverage = ", round(mean(test.intervals$covered) * 100, 1), "%")

p.conformal <- ggplot(test.intervals) +
  geom_pointrange(aes(x = observed,
                      y = .pred,
                      ymin = .pred_lower,
                      ymax = .pred_upper,
                      color = covered),
                  alpha = 0.6, linewidth = 0.4) +
  geom_abline(intercept = 0, slope = 1) +
  annotate(geom = "text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.2,
            label = conformal.annotation, size = 3) +
  scale_color_manual(values = c("TRUE" = "steelblue", "FALSE" = "tomato")) +
  labs(title = "Soil Organic Carbon (wt%) and prediction intervals (PI95%)",
       x = "Observed", y = "Predicted",
       color = "PI95% covered:") +
  theme_light() +
  theme(legend.position = "bottom")

r.max <- max(layer_scales(p.conformal)$x$range$range)
r.min <- min(layer_scales(p.conformal)$x$range$range)
s.max <- max(layer_scales(p.conformal)$y$range$range)
s.min <- min(layer_scales(p.conformal)$y$range$range)
t.max <- round(max(r.max, s.max), 1)
t.min <- round(min(r.min, s.min), 1)

p.conformal + coord_equal(xlim = c(t.min, t.max), ylim = c(t.min, t.max))
```
